---
title: "Piecewise Linear Regression for Groundwater Trends"
author: "Luna Herschenfeld-Catalan"
---

# Data Access

## Groundwater Data

Access groundwater data [here](https://zenodo.org/records/10003697). 

Citation: Groundwater level data, aquifer system boundaries, and Supplementary Tables associated with Jasechko, S., Seybold, H., Perrone, D., Fan, Y., Shamsudduha, M., Taylor, R.G., Fallatah, O., Kirchner, J.W. Rapid groundwater decline and some cases of recovery in aquifers globally. Nature, https://doi.org/10.1038/s41586-023-06879-8 (2024).

- Depth to groundwater is saved as `AnnualDepthToGroundwater.csv`. Due to the way its measured, positive slope values signify declines in groundwater. 
- Groundwater elevation is saved as `AnnualGroundwaterElevation.csv`. 

## US data

Accessed us boundary data [here](https://www.census.gov/cgi-bin/geo/shapefiles/index.php?year=2023&layergroup=States+%28and+equivalent%29)

# Data Exploration 

```{r setup, include=FALSE, message=FALSE, warning = FALSE}
library(tidyverse)
library(here)

# to make geospatil objects
library(sf)

library(usmap)

library(forecast)

# fill in the missing NA values
library(performanceEstimation)

library(tidyfit)
library(lubridate)
```


Analysis should include 
```{r message=FALSE, warning = FALSE}
# read in the depth data
depth <- read_csv(here('data', "AnnualDepthToGroundwater.csv"))

# read in the elevation data
elevation <- read_csv(here('data', "AnnualGroundwaterElevation.csv"))

# reda in map data
us <- st_read(here("data", "tl_2023_us_state", "tl_2023_us_state.shp")) %>% 
  janitor::clean_names() %>% 
  select(name, geometry)
```


```{r message=FALSE, warning = FALSE}

location <- depth %>% 
  select(StnID, Lat, Lon) %>% 
  distinct() %>% # select the unique well names
  st_as_sf(coords = c("Lon", "Lat")) # set the geographic coordinates

# set coordinate reference system
location <- location %>% 
  st_set_crs(st_crs(us)) # same crs as us df

station_counts <- depth %>% 
  group_by(StnID) %>% 
  summarize(count = n()) %>% 
  filter(count > 10)

depth_clean <- depth %>% 
  filter(StnID %in% station_counts$StnID)

# check
##st_crs(location)

# all the points - this isnt super helpful 
# ggplot(location) +
#   geom_sf(size = 0.5)
  
```


## Create a subset of Texas Wells

```{r message=FALSE, warning = FALSE}
# filter for texas
texas <- us %>% 
  filter(name == "Texas")

# plot(texas)

# check
##st_crstexas) == st_crs(location)

# select the well locations within Texas
texas_wells <- location %>% 
  st_intersection(texas)

# select the texas wells that have more than 10 observations
texas_wells_count <- texas_wells %>% 
  filter(StnID %in% station_counts$StnID) # select IDs that match the Texas location

# ggplot(texas_wells) +
#   geom_sf(size = 0.5)
# 
# ggplot(texas_wells_count) +
#   geom_sf(size = 0.5)
```

Select all the well data in Texas
- What is the time buffer that I want for the wells? **10 years**

```{r message=FALSE, warning = FALSE}

depth_texas <- depth %>% 
  filter(StnID %in% texas_wells_count$StnID) # select IDs that match the Texas location

  
```



```{r warning = FALSE, message = FALSE}
subset <- depth_texas %>% 
  mutate(year = IntegerYear) %>% 
  filter(StnID == "1040141399") %>% 
  arrange(year)

# generate a time sequence to fill in missing dates
subset_years <- data.frame(year = seq(min(subset$year), max(subset$year)))

# merge years with dataframe
subset <- merge(subset, subset_years, all = T) %>% 
  mutate(valley_flag = ifelse(lag(year, 5) %in% year & lead(year, 5) %in% year, TRUE, FALSE))

ggplot() +
  geom_line(data = subset, 
             aes(x = year, y = DepthToWater_m)) +
  scale_y_reverse()
```


```{r eval = FALSE, warning = FALSE, message = FALSE}

valleys_df <- subset %>%
  mutate(valley_flag = ifelse(valley_flag == TRUE, 
                              ifelse(DepthToWater_m <= lag(DepthToWater_m, 5) & DepthToWater_m <= lead(DepthToWater_m, 5), 
                                     "valley", 
                                     "not"), 
                              NA))

ggplot() +
  geom_point(data = valleys_df, 
             aes(x = year, y = DepthToWater_m, color = valley_flag)) +
  scale_y_reverse()

```

## Create Rolling Window Regression 

Create a rolling window and calculate a linear regression each 4 years: 

```{r warning = FALSE, message = FALSE, eval = FALSE}
results <- subset %>%
    mutate(rolling_index = row_number()) %>%
    group_by(rolling_index)

index <- first(results$rolling_index)
start_date <- 1958#results$year[index]
end_year <- start_date + 5
window_data <- results %>%
     filter(year >= start_date & year <= end_year)
lm_results <- lm(DepthToWater_m ~ year, data = window_data)
slope <- summary(lm_results)$coefficients[2] # get the slope
data.frame(start_date = start_date, end_date = end_date, coefficients)


start_date <- 2000
end_year <- 2005
window_data <- results %>%
  filter(year >= start_date & year <= end_year)
lm_results <- lm(DepthToWater_m ~ year, data = window_data)
summary(lm_results)
slope <- summary(lm_results)$coefficients[2] # get the slope
df <- data.frame(start_date = start_date, 
                 end_date = end_year, 
                 slope = slope) 
  
  coef_df <- rbind(coef_df, df)

results <- subset %>%
    mutate(rolling_index = row_number()) %>%
    group_by(rolling_index)

coef_df <- data.frame()

years <- c(results$year)

for(start_year in 1:nrow(results)) {
  
  start_date <- years[start_year]
  end_year <- start_date + 4
  
  window_data <- results %>%
     filter(year >= start_date & year <= end_year)
  
  lm_results <- lm(DepthToWater_m ~ year, 
                   data = window_data)
  slope <- summary(lm_results)$coefficients[2] # get the slope
  df <- data.frame(start_date = start_date, 
                   end_date = end_year, 
                   slope = slope) 
  
  coef_df <- rbind(coef_df, df)
  
}

sub <- data.frame()

seq_yr <- seq(1, nrow(coef_df), by = 4)

for(i in seq(seq_yr)) {
  
  sub <- coef_df[i]  # Select every 5 rows
  print(sub)
  
}

years <- c(results$year)

for(start_year in seq_along(years)) {
  #print(start_year)
  
  start_date <- years[start_year]
  #print(start_date)
  end_year <- start_date + 5
  ##print(end_year)
  window_data <- results %>%
    filter(year >= start_date & year <= end_year)
  
  # Check if there are any NA values in DepthToWater_m
  if (any(is.na(window_data$DepthToWater_m))) {
    next
  }
  
  lm_results <- lm(DepthToWater_m ~ year, 
                   data = window_data)
  slope <- summary(lm_results)$coefficients[2] # get the slope
  df <- data.frame(start_date = start_date, 
                   end_date = end_year, 
                   slope = slope) 
  
  coef_df <- rbind(coef_df, df)
}

```

```{r}

subset <- subset %>% 
  mutate(index = row_number())

coef_df <- data.frame()
  
for(position in seq(1, nrow(subset), by = 5)) {
    
    start_position <- position
    #print(start_position)

    end_position <- start_position + 5
    #print(end_position)

    window_data <- subset %>%
      filter(index >= start_position & index <= end_position)
    
    #print(length(window_data))
    
    # Check if there are any NA values in DepthToWater_m
    # if (any(is.na(window_data$DepthToWater_m))) {
    #   next
    # }
    
    lm_results <- lm(DepthToWater_m ~ IntegerYear, 
                     data = window_data)
    
    slope <- summary(lm_results)$coefficients[2] # get the slope
    print(slope)
    
    df <- data.frame(index = position,
                     start_date = start_position, 
                     end_date = end_position, 
                     slope = slope) 
    
    coef_df <- rbind(coef_df, df)
    
    
}

# I dont know whats going o and why this is not doing what I want it to 
#window_data

#yay$coefficients
```


# Create function for rolling regression

```{r warning = FALSE, message = FALSE}
# Function to calculate rolling linear regression
rolling_lm <- function(data, window_size) {
  
  data <- data %>% 
    mutate(index = row_number())
  
  coef_df <- data.frame()
  
  for(position in seq(1, nrow(data), by = window_size)) {
    
    start_position <- position

    end_position <- start_position + window_size

    window_data <- data %>%
      filter(index >= start_position & index <= end_position)
    
    # Check if there are any NA values in DepthToWater_m
    # if (any(is.na(window_data$DepthToWater_m))) {
    #   next
    # }
    
    lm_results <- lm(DepthToWater_m ~ index, 
                     data = window_data)
    
    slope <- summary(lm_results)$coefficients[2] # get the slope
    
    df <- data.frame(index = position,
                     start_date = data$year[start_position], 
                     end_date = data$year[end_position], 
                     slope = slope) 
    
    coef_df <- rbind(coef_df, df)
  }
  
  return(coef_df)
}
```


Want code to be able to detect a case like saudis arabia 
    - Change in sign of slope (if using depth to water: positive to negative slope)
    - Change in magnitude (product of slopes should be negative; difference between change in slopes slopes should be > 0.5m/yr)
    
```{r warning = FALSE, message = FALSE}

source("funs/rolling_lm_index.R")
source("funs/rolling_lm_yr.R")

# Calculate rolling linear regression with a 5-year window
rolling_results <- rolling_lm_yr(subset, window_size = 5) %>% 
  
  # calculate the change in slope
  mutate(change = c(NA, diff(slope)),
         
         # calculate the sign change
         sign_change = c(0, diff(sign(slope)))) %>%  
  
  # identify when the magnitude change is greater than 0.5 and the sign changes from pos to neg
  mutate(flag = ifelse(abs(change) >= 0.5 & sign_change == -2,
                       "breakpoint",
                       NA)) %>% 
  filter(flag == "breakpoint")

```


```{r}

ggplot(data = subset, 
             aes(x = year, y = DepthToWater_m)) +
  geom_point() +
  geom_vline(data = rolling_results, 
             aes(xintercept = start_date), 
             color = "red",  # Adjust color as needed
             linetype = "dashed") +  # Adjust linetype as needed
  scale_y_reverse()

```

# Loading in the Houston data

```{r}
houston <- readxl::read_xlsx(here::here("data", "houston_gw.xlsx")) %>% 
  select(Date, DepthToWater_m = DepthToGW_m) %>% 
  mutate(year = stringr::str_extract(Date, ".*(?=\\.)")) %>% 
  mutate(year = as.numeric(year)) %>% 
  group_by(year) %>% 
  summarize(DepthToWater_m = mean(DepthToWater_m)) %>% 
  ungroup()

houston_lm <- rolling_lm_index(houston, window_size = 4) %>% 
  
  # calculate the change in slope
  mutate(change = c(NA, diff(slope)),
         
         # calculate the sign change
         sign_change = c(0, diff(sign(slope))),
         
         start_date = as.numeric(start_date),
         end_date = as.numeric(end_date)) %>%  
  
  # identify when the magnitude change is greater than 0.5 and the sign changes from pos to neg
  mutate(flag = ifelse(abs(change) >= 0.5 & sign_change == -2,
                       "breakpoint",
                       NA)) %>% 
  filter(flag == "breakpoint")

```

## Create nested functions to identify breakpoints

```{r}

breakpoint_fun <- rolling_lm_index <- function(data, window_size) {
  
  data <- data %>% 
    mutate(index = row_number())
  
  coef_df <- data.frame()
  
  for(position in seq(1, nrow(data), by = window_size)) {
    
    start_position <- position
    
    end_position <- start_position + window_size
    
    window_data <- data %>%
      filter(index >= start_position & index <= end_position)
    
    # Check if there are any NA values in DepthToWater_m
    # if (any(is.na(window_data$DepthToWater_m))) {
    #   next
    # }

    lm_results <- lm(DepthToWater_m ~ index, 
                     data = window_data)
    
    slope <- summary(lm_results)$coefficients[2] # get the slope
    
    df <- data.frame(index = position,
                     start_date = data$year[position], 
                     end_date = data$year[end_position], 
                     slope = slope) 
    
    coef_df <- rbind(coef_df, df)
  }
  
  breakpoint_df <- coef_df %>% 
    # calculate the change in slope
    mutate(change = c(NA, diff(slope)),
           
           # calculate the sign change
           sign_change = c(0, diff(sign(slope))),
           
           start_date = as.numeric(start_date),
           end_date = as.numeric(end_date)) %>%  
    
    # identify when the magnitude change is greater than 0.5 and the sign changes from pos to neg
    mutate(flag = ifelse(abs(change) >= 0.5 & sign_change == -2,
                         "breakpoint",
                         NA)) %>% 
    filter(flag == "breakpoint")
  
  return(breakpoint_df)
}

```
### Test on the Houston Data

```{r}
houston_lm <- breakpoint_fun(houston, window_size = 5)

ggplot() +
  geom_point(data = houston, 
             aes(x = year, y = DepthToWater_m)) +
  geom_vline(data = houston_lm, 
             aes(xintercept = start_date), 
             color = "red",  # Adjust color as needed
             linetype = "dashed") +  # Adjust linetype as needed
  scale_y_reverse()
```

# Apply the function to the whole `depth_texas`

I encountered a problem with a datatype like the one below. How do I deal with these? 

```{r}
depth_id <- depth_texas %>% 
  select(StnID) %>% 
  group_by(StnID) %>%
  distinct() %>% 
  head(5)

# depth_id[[1]][1]
# 
subset <- depth_texas %>%
    filter(StnID %in% depth_id[[1]][1])  %>% 
    mutate(year = IntegerYear)

# this one looks at the years
subset_lm_yr2 <- rolling_lm_yr(subset, window_size = 4)
subset_lm_index2 <- rolling_lm_index(subset, window_size = 4)

# no breakpoint identified 
subset_bp <- breakpoint_fun(subset, window_size = 4)

ggplot() +
  geom_point(data = subset, 
             aes(x = year, y = DepthToWater_m)) +
  scale_y_reverse()
```


Troubleshooting the above code that returns 0 observations
- The code works, the `rolling_lm_index()` does not detect any changes. However, the `rolling_lm_yr` does product observations. 
- Okay just kidding, now the functions work

Next steps: 

1. Run the function, and if 0 breakpoint observations are returned then store that `StnID` but continue going. Then, we can investigate the `StnID`'s that are in that dataframe to see if they are worse case scenarios. 

```{r}
# select 5 IDs
depth_id <- depth_texas %>% 
  select(StnID) %>% 
  group_by(StnID) %>%
  distinct() %>% 
  head(5)

# create subset of data with 5 wells
subset <- depth_texas %>%
    filter(StnID %in% depth_id$StnID)  %>% 
    mutate(year = IntegerYear)

# check that it subsetted correctly
unique(depth_id$StnID) == unique(subset$StnID)
```

What do I want the function to do?
- Take in a complete data set
- take the unique groups (use distinct to make a list?)
- iterate through that list and filter for only that data 
- run the breakpoint function on it
- if the nrow(sub_breakpoint) < 1 save the StnID to a list 
- if the nrow(subset_bp) > 1 save the StnID to a dataframe

Framework: 
```{r}
# create list of unique IDs
id_list <- subset$StnID %>% 
  unique()

# filter for only a single station
sub <- subset %>% 
  filter(StnID == id_list[3])

# run the breakpoint function on it
bp_df <- breakpoint_fun(sub, window_size = 5)

no_bp_list <- list()
bp_list <- list()

if (nrow(bp_df) < 1) {
  no_bp_list[2] <- sub$StnID
}

if (nrow(bp_df) >= 1) {
  bp_list[2] <- sub$StnID
}
```

## Create the for loop to iterate over the id_list
```{r}

no_bp <- data.frame()
bp <- data.frame()

name <- "StnID"


for (id in seq(id_list)) {
  
  # filter for only a single station
  sub <- subset %>% 
    filter(StnID == id_list[id])
  
  ID = id_list[id]
  
  # run the breakpoint function on it
  bp_df <- breakpoint_fun(sub, window_size = 5)
  
  ifelse(nrow(bp_df) < 1,
         no_bp <- rbind(no_bp, ID),
         bp <- rbind(bp, ID))
  
}

colnames(no_bp) <- name
colnames(bp) <- name
```

Check the output
- It looks like its saving the right ones!
```{r}
id_list[5]
# filter for only a single station
sub <- subset %>% 
  filter(StnID == id_list[5])

# run the breakpoint function on it
bp_df <- breakpoint_fun(sub, window_size = 5)

ggplot() +
  geom_point(data = sub, 
             aes(x = year, y = DepthToWater_m)) +
  geom_vline(data = bp_df, 
             aes(xintercept = start_date), 
             color = "red",  # Adjust color as needed
             linetype = "dashed") +  # Adjust linetype as needed
  theme_minimal() +
  scale_y_reverse()
```
# Create function that takes in dataframe and assigns them as breakpoints or not

```{r}

breakpoint_id <- function(df) {
  
  # create list of unique IDs
  id_list <- df$StnID %>% 
    unique()
  
  
  no_bp <- data.frame()
  bp <- data.frame()
  
  name <- "StnID"
  
  
  for (id in seq(id_list)) {
    
    # filter for only a single station
    sub <- df %>% 
      filter(StnID == id_list[id])
    
    ID = id_list[id]
    
    # run the breakpoint function on it
    bp_df <- breakpoint_fun(sub, window_size = 5)
    
    ifelse(nrow(bp_df) < 1,
           no_bp <- rbind(no_bp, ID),
           bp <- rbind(bp, ID))
    
    # plot <- ggplot() +
    #   geom_point(data = sub, 
    #              aes(x = year, y = DepthToWater_m)) +
    #   geom_vline(data = bp_df, 
    #              aes(xintercept = start_date), 
    #              color = "red",  # Adjust color as needed
    #              linetype = "dashed") +  # Adjust linetype as needed
    #   theme_classic() +
    #   labs(y = "Depth To Water (m)",
    #        x = "Year") +
    #   scale_y_reverse()
    # 
    # path = '/Users/lunacatalan/Documents/dev/Jasechko/groundwater-trends/plots'
    # file_name = paste("plot_", ID, ".png", sep="")
    # 
    # ggsave(path = path, filename = file_name, width = 10, height = 5, dpi = 150, units = "in")
    
  }
  
  colnames(no_bp) <- name
  colnames(bp) <- name
  
  # return a list of dataframe
  return(list(bp = bp, non = no_bp))
  
}

```


## Test the function
```{r}

ids_test <- breakpoint_id(subset)

ids_test$bp
ids_test$non
```

# Apply the function on the texas data subset
```{r warning = FALSE, message=FALSE, eval = FALSE}
source("funs/breakpoint_fun.R")

depth_texas <- depth_texas %>% 
  mutate(year = IntegerYear)

# run the function
system.time(
  ids <- breakpoint_id(depth_texas)
)

# extract the wells with breakpoints and non
breakpoint_wells <- ids$bp
non_wells <- ids$non

# select the geometries of the points with breakpoints labelled 
depth_texas_id <- texas_wells_count %>% 
  mutate(breakpoint = ifelse(texas_wells_count$StnID %in% breakpoint_wells$StnID,
                             "yes",
                             "no"))
# filter for places where there is a breakpoint
breakpoint <- depth_texas_id %>% 
  filter(breakpoint == "yes")
# filter for places with no breakpoint
non <- depth_texas_id %>% 
  filter(breakpoint == "no")

# plot the wells
plot <- ggplot() +
  geom_sf(data = non, size = 0.5, color = "darkgray") +
  geom_sf(data = breakpoint, size = 0.5, color = "blue", alpha = 0.5) + 
  theme_minimal()


setwd("plots") #where the .docxs live
news_files <- list.files(pattern = ".png", 
                         path = getwd(),
                         full.names = TRUE, recursive = TRUE, ignore.case = TRUE)
```


Things I learned
- Some of the plots with no breakpoints have been experiencing shallowing of groundwater the entire time frame, so there was no change in slope. To counter this: if the sign is negative 3 times in a row --> flag it
- Some of the plots are declining but did experience a slow in the decline. To counter this: if there is decrease in the slope but not in sign 


