---
title: "Piecewise Linear Regression for Groundwater Trends"
author: "Luna Herschenfeld-Catalan"
---

# Data Access

## Groudnwater Data

Access groundwater data [here](https://zenodo.org/records/10003697). 

Citation: Groundwater level data, aquifer system boundaries, and Supplementary Tables associated with Jasechko, S., Seybold, H., Perrone, D., Fan, Y., Shamsudduha, M., Taylor, R.G., Fallatah, O., Kirchner, J.W. Rapid groundwater decline and some cases of recovery in aquifers globally. Nature, https://doi.org/10.1038/s41586-023-06879-8 (2024).

- Depth to groundwater is saved as `AnnualDepthToGroundwater.csv`. Due to the way its measured, positive slope values signify declines in groundwater. 
- Groundwater elevation is saved as `AnnualGroundwaterElevation.csv`. 

## US data

Accessed us boundary data [here](https://www.census.gov/cgi-bin/geo/shapefiles/index.php?year=2023&layergroup=States+%28and+equivalent%29)

# Data Exploration 

```{r setup, include=FALSE, message=FALSE, warning = FALSE}
library(tidyverse)
library(here)

# to make geospatil objects
library(sf)

library(usmap)
```


Analysis should include 
```{r message=FALSE, warning = FALSE}
# read in the depth data
depth <- read_csv(here('data', "AnnualDepthToGroundwater.csv"))

# read in the elevation data
elevation <- read_csv(here('data', "AnnualGroundwaterElevation.csv"))

# reda in map data
us <- st_read(here("data", "tl_2023_us_state", "tl_2023_us_state.shp")) %>% 
  janitor::clean_names() %>% 
  select(name, geometry)
```


```{r message=FALSE, warning = FALSE}

location <- depth %>% 
  select(StnID, Lat, Lon) %>% 
  distinct() %>% # select the unique well names
  st_as_sf(coords = c("Lon", "Lat")) # set the geographic coordinates

# set coordinate reference system
location <- location %>% 
  st_set_crs(st_crs(us)) # same crs as us df

# check
##st_crs(location)

# all the points - this isnt super helpful 
ggplot(location) +
  geom_sf(size = 0.5)
  
```


Create a subset of Texas Wells

```{r message=FALSE, warning = FALSE}
# filter for texas
texas <- us %>% 
  filter(name == "Texas")

plot(texas)

# check
##st_crstexas) == st_crs(location)

# select the well locations within Texas
texas_wells <- location %>% 
  st_intersection(texas)

ggplot(texas_wells) +
  geom_sf(size = 0.5)
```

Select all the well data in Texas
- What is the time buffer that I want for the wells? 
- Do I only want monitoring wells that span 40+ years?

```{r message=FALSE, warning = FALSE}

depth_texas <- depth %>% 
  filter(StnID %in% texas_wells$StnID) # select IDs that match the Texas location
  
```
