---
title: "Piecewise Linear Regression for Groundwater Trends"
author: "Luna Herschenfeld-Catalan"
---

# Data Access

## Groundwater Data

Access groundwater data [here](https://zenodo.org/records/10003697). 

Citation: Groundwater level data, aquifer system boundaries, and Supplementary Tables associated with Jasechko, S., Seybold, H., Perrone, D., Fan, Y., Shamsudduha, M., Taylor, R.G., Fallatah, O., Kirchner, J.W. Rapid groundwater decline and some cases of recovery in aquifers globally. Nature, https://doi.org/10.1038/s41586-023-06879-8 (2024).

- Depth to groundwater is saved as `AnnualDepthToGroundwater.csv`. Due to the way its measured, positive slope values signify declines in groundwater. 
- Groundwater elevation is saved as `AnnualGroundwaterElevation.csv`. 

## US data

Accessed us boundary data [here](https://www.census.gov/cgi-bin/geo/shapefiles/index.php?year=2023&layergroup=States+%28and+equivalent%29)

# Data Exploration 

```{r setup, include=FALSE, message=FALSE, warning = FALSE}
library(tidyverse)
library(here)

# to make geospatil objects
library(sf)

library(usmap)

library(forecast)

# fill in the missing NA values
library(performanceEstimation)

library(tidyfit)
library(lubridate)

library(purrr)

# for map
library(gganimate)
library(gifski)
# for interactive mao
library(plotly)
library(htmlwidgets)
library(leaflet)
```

Read in functions
```{r}
# read in the functions
source(here::here("funs/rolling_lm_index.R"))
source(here::here("funs/rolling_lm_yr.R"))
source(here::here("funs/breakpoint_fun.R"))
source(here::here("funs/breakpoint_id.R"))
```

Analysis should include 
```{r message=FALSE, warning = FALSE}
# read in the depth data
depth <- read_csv(here('data', "AnnualDepthToGroundwater.csv"))

# read in the elevation data
elevation <- read_csv(here('data', "AnnualGroundwaterElevation.csv"))

# reda in map data
us <- st_read(here("data", "tl_2023_us_state", "tl_2023_us_state.shp")) %>% 
  janitor::clean_names() %>% 
  select(name, geometry)


continental_us <- us %>% 
  filter(!(name %in% c("American Samoa", 
                       "Alaska", 
                       "Puerto Rico", 
                       "Commonwealth of the Northern Mariana Islands", 
                       "Hawaii", 
                       "United States Virgin Islands"))) %>% 
  st_crop(worldmap, xmin = -125.0, xmax = -66.93457,
                                    ymin = 24.396308, ymax = 49.384358)
```

Select wells with more than 10 observations, and distinct geometries for each well. 
```{r message=FALSE, warning = FALSE}

station_counts <- depth %>% 
  group_by(StnID) %>% 
  summarize(count = n()) %>% 
  filter(count > 10)

# Wells with data that occur more than 10 times
depth_clean <- depth %>% 
  filter(StnID %in% station_counts$StnID)

# location of the wells that have 10+ observations
location <- depth_clean %>% 
  select(StnID, Lat, Lon) %>% 
  distinct() %>% # select the unique well names
  st_as_sf(coords = c("Lon", "Lat")) # set the geographic coordinates

# set coordinate reference system
location <- location %>% 
  st_set_crs(st_crs(us)) # same crs as us df


# check
##st_crs(location)
```

## Create a subset of Texas Wells

```{r message=FALSE, warning = FALSE}
# filter for texas
texas <- us %>% 
  filter(name == "Texas")

# check
##st_crstexas) == st_crs(location)

# select the well locations within Texas that have more than 10 observations
texas_wells <- location %>% 
  st_intersection(texas)

# select all the wells in texas
depth_texas <- depth %>% 
  filter(StnID %in% texas_wells$StnID) # select IDs that match the Texas location
```

# Loading in the Houston data
```{r message=FALSE, warning = FALSE}
# load in houston data to check functionality
houston <- readxl::read_xlsx(here::here("data", "houston_gw.xlsx")) %>% 
  select(Date, DepthToWater_m = DepthToGW_m) %>% 
  mutate(year = stringr::str_extract(Date, ".*(?=\\.)")) %>% 
  mutate(year = as.numeric(year)) %>% 
  group_by(year) %>% 
  summarize(DepthToWater_m = mean(DepthToWater_m)) %>% 
  ungroup()
```

Run the regression on the data 
```{r message=FALSE, warning = FALSE}

houston_lm <- rolling_lm_index(houston, window_size = 4) %>% 
  
  # calculate the change in slope
  mutate(change = c(NA, diff(slope)),
         
         # calculate the sign change
         sign_change = c(0, diff(sign(slope))),
         
         start_date = as.numeric(start_date),
         end_date = as.numeric(end_date)) %>%  
  
  # identify when the magnitude change is greater than 0.5 and the sign changes from pos to neg
  mutate(flag = ifelse(abs(change) >= 0.5 & sign_change == -2,
                       "breakpoint",
                       NA)) %>% 
  filter(flag == "breakpoint")
```

Test the breakpoint_fun on the houston data
```{r message=FALSE, warning = FALSE}
### Test on the Houston Data
houston_lm <- breakpoint_fun(houston, window_size = 5)

ggplot() +
  geom_point(data = houston, 
             aes(x = year, y = DepthToWater_m)) +
  geom_vline(data = houston_lm, 
             aes(xintercept = start_date), 
             color = "red",  # Adjust color as needed
             linetype = "dashed") +  # Adjust linetype as needed
  scale_y_reverse()
```

# Label the Texas well breakpoints
```{r message=FALSE, warning = FALSE}
# prepare the texas data for processing
depth_texas <- depth_texas %>% 
  mutate(year = IntegerYear)

# run the id function
system.time(
  texas_id <- breakpoint_id(depth_texas)
)

# check to see if all the ids were looked at - should be TRUE
length(unique(texas_id$bp$StnID)) + length(unique(texas_id$non$StnID)) == length(unique(depth_texas$StnID))

# Define a function to generate the sequence of years
generate_years <- function(start_date, end_date) {
  seq(start_date, end_date, by = 1)
}

texas_breakpoints <- texas_id$bp %>% 
  select(StnID, start_date, end_date) %>% 
  # the number of years that this extends
  mutate(years = end_date - start_date + 1) %>% 
  mutate(years = ifelse(is.na(years),
                        1,
                        years),
         end_date = ifelse(is.na(end_date),
                           start_date,
                           end_date)) %>% 
  #uncount(years, .remove = FALSE) %>% 
  group_by(StnID) %>%
  mutate(range = map2(start_date, end_date, generate_years)) %>% 
  unnest(cols = range) %>% 
  select(StnID, range) %>% 
  arrange(range) %>% 
  ungroup()

texas_year_plot <- texas_wells %>% 
  select(-name) %>% 
  inner_join(texas_breakpoints, by = "StnID")

# replace this with desired CRS
texas_year_plot <- st_transform(texas_year_plot, crs = "+init=EPSG:4326")


# extract the wells with breakpoints and non
breakpoint_wells <- texas_id$bp
non_wells <- texas_id$non

# select the geometries of the points with breakpoints labelled 
depth_texas_id <- texas_wells %>% 
  mutate(breakpoint = ifelse(texas_wells$StnID %in% breakpoint_wells$StnID,
                             "yes",
                             "no"))
# filter for places with no breakpoint
non <- depth_texas_id %>% 
  filter(breakpoint == "no")
```

Create animated map of the breakpoints over time
```{r message=FALSE, warning = FALSE}
texas_time <- ggplot() +
  # plot the outline of texas
  geom_sf(data = texas, fill = "white") +
  # plot the wells with no breakpoints
  geom_sf(data = non, size = 0.5, color = "grey", alpha = 0.5) + 
  # plot the wells with breakpoints
  geom_sf(data = texas_year_plot, size = 0.5, color = "blue", alpha = 0.5) + 
  theme_minimal()

map_with_animation <- texas_time +
  transition_time(range) +
  ggtitle('Year: {frame_time}',
          subtitle = 'Frame {frame} of {nframes}')
num_years <- max(texas_year_plot$range) - min(texas_year_plot$range) + 1

animated_map <- animate(map_with_animation, 
                        nframes = num_years, 
                        renderer = gifski_renderer())

anim_save("animated_texas.gif")

```

Next steps:
- Colorcode breakpoints by year
- Don't need specific wells --> need locations 
- Sequence of maps that describe time intervals 

Use the `texas_year_plot` dataframe that has the list of wells that have breakpoints, and the years that they occur. 
- Split into 10 year intervals. The years go from 1934-2022
```{r}
# goes from 
range(texas_year_plot$range)

# create bins of 10 years
texas_year_bins <- texas_year_plot %>% 
  mutate(bin = case_when(range %in% c(1934:1944) ~ "1934-1944", 
                         range %in% c(1945:1955) ~ "1945-1955", 
                         range %in% c(1956:1966) ~ "1956-1966", 
                         range %in% c(1967:1977) ~ "1967-1977",
                         range %in% c(1978:1988) ~ "1978-1988", 
                         range %in% c(1989:1999) ~ "1989-1999", 
                         range %in% c(2000:2010) ~ "2000-2010",
                         range %in% c(2011:2022) ~ "2011-2022")) %>% 
  select(!range) %>% 
  st_drop_geometry() %>% 
  pivot_wider(names_from = "bin",
              values_from = "StnID")


# select the range
plot_1944 <- texas_year_bins["1934-1944"] %>% 
  # unnest the list of station IDs
  unnest(cols = c(`1934-1944`)) %>% 
  rename(StnID = `1934-1944`) %>% 
  # select only the distinct StnIDs
  distinct() %>% 
  left_join(texas_wells) %>%
  select(!name) %>% 
  st_as_sf()

ggplot() +
  # plot the outline of texas
  geom_sf(data = texas, fill = "white") +
  # plot the wells with no breakpoints
  geom_sf(data = non, size = 0.5, color = "grey", alpha = 0.5) + 
  # plot the wells with breakpoints for that range
  geom_sf(data = plot_1944, size = 0.5, color = "blue", alpha = 0.5) + 
  theme_minimal()
```

Create plots for every range of values
- Map of year interval defined above. Here there are 8 10 year intervals (with the exception of 2011-2022 because I wanted to make sure all the data was represented)
- All of the wells are represented in gray on the map, and only the wells that experienced a breakpoint during that 10 year interval are highlighted. 
- plots saved to `plots` with range defined in image name (eg. plot_1989-1999 is a plot of the wells in Texas that had a breakpoint between 1989 and 1999)

```{r}

for (range in seq(texas_year_bins)) {
  
  range_df <- texas_year_bins[range]
  col_name <- colnames(range_df)
  
  plot_data <- range_df %>% 
    # unnest the list of station IDs
    unnest(cols = c(col_name)) %>% 
    rename(StnID = col_name) %>% 
    # select only the distinct StnIDs
    distinct() %>% 
    left_join(texas_wells) %>%
    select(!name) %>% 
    st_as_sf()
    
  plot <- ggplot() +
    # plot the outline of texas
    geom_sf(data = texas, fill = "white") +
    # plot all the wells in texas
    geom_sf(data = texas_wells, size = 0.05, color = "grey", alpha = 0.5) + 
    # plot the wells with breakpoints for that range
    geom_sf(data = plot_data, size = 0.05, color = "blue", alpha = 0.5) + 
    theme_classic()
  
  path = '/Users/lunacatalan/Documents/dev/Jasechko/groundwater-trends/plots'
  file_name = paste("plot_", col_name, ".png", sep="")
  ggsave(path = path, filename = file_name, width = 10, height = 5, dpi = 150, units = "in")
  
}

```


This is the set up to select the wells that are within the continental US. Because of computation time, this did not work on my computer.

- The acronym `cus` refers to `continental US`
```{r}
# select the well locations within continental US (cus) that have more than 10 observations
system.time(
  cus_wells <- location %>% 
    st_intersection(continental_us)
)

# select all the wells in the cus
cus_depth <- depth %>% 
  filter(StnID %in% cus_wells$StnID) # select IDs that match the Texas location

# prepare the cus data for processing
cus_depth <- cus_depth %>% 
  mutate(year = IntegerYear)

# run the id function
system.time(
  cus_id <- breakpoint_id(cus_depth)
)

# check to see if all the ids were looked at - should be TRUE
length(unique(cus_id$bp$StnID)) + length(unique(cus_id$non$StnID)) == length(unique(cus_depth$StnID))
```

If the above is true, then move on to the next code chunk. 

```{r}
# Define a function to generate the sequence of years
generate_years <- function(start_date, end_date) {
  seq(start_date, end_date, by = 1)
}

cus_breakpoints <- cus_id$bp %>% 
  select(StnID, start_date, end_date) %>% 
  # the number of years that this extends
  mutate(years = end_date - start_date + 1) %>% 
  mutate(years = ifelse(is.na(years),
                        1,
                        years),
         end_date = ifelse(is.na(end_date),
                           start_date,
                           end_date)) %>% 
  #uncount(years, .remove = FALSE) %>% 
  group_by(StnID) %>%
  mutate(range = map2(start_date, end_date, generate_years)) %>% 
  unnest(cols = range) %>% 
  select(StnID, range) %>% 
  arrange(range) %>% 
  ungroup()

cus_year_plot <- cus_wells %>% 
  select(-name) %>% 
  inner_join(cus_breakpoints, by = "StnID")

# replace this with desired CRS -- is this the right crs?
cus_year_plot <- st_transform(cus_year_plot, crs = "+init=EPSG:4326")


# extract the wells with breakpoints and non
breakpoint_wells <- cus_id$bp
non_wells <- cus_id$non

# select the geometries of the points with breakpoints labelled 
cus_depth_id <- cus_wells %>% 
  mutate(breakpoint = ifelse(cus_wells$StnID %in% breakpoint_wells$StnID,
                             "yes",
                             "no"))
# filter for places with no breakpoint
non <- cus_depth_id %>% 
  filter(breakpoint == "no")
```

Map the intervals for cus data
- edit the ranges

```{r, message=FALSE, warning=FALSE}
# goes from 1865 2022
range(depth_clean$IntegerYear)

# create bins of 10 years
cus_year_bins <- cus_year_plot %>% 
  mutate(bin = case_when(range %in% c(1865:1875) ~ "1865-1875", 
                         range %in% c(1876:1886) ~ "1876-1886",
                         range %in% c(1887:1897) ~ "1887-1897",
                         range %in% c(1898:1908) ~ "1898-1908",
                         range %in% c(1909:1919) ~ "1909-1919",
                         range %in% c(1920:1930) ~ "1920-1930",
                         range %in% c(1931:1941) ~ "1931-1941",
                         range %in% c(1942:1952) ~ "1942-1952",
                         range %in% c(1953:1963) ~ "1953-1963", 
                         range %in% c(1964:1974) ~ "1964-1974",
                         range %in% c(1975:1985) ~ "1975-1985", 
                         range %in% c(1986:1996) ~ "1986-1996", 
                         range %in% c(1997:2007) ~ "1997-2007",
                         range %in% c(2008:2018) ~ "2008-2018",
                         range %in% c(2019:2022) ~ "2019-2022")) %>% 
  select(!range) %>% 
  st_drop_geometry() %>% 
  pivot_wider(names_from = "bin",
              values_from = "StnID")
```
Save still plots of maps:

```{r}

for (range in seq(cus_year_bins)) {
  
  range_df <- cus_year_bins[range]
  col_name <- colnames(range_df)
  
  plot_data <- range_df %>% 
    # unnest the list of station IDs
    unnest(cols = c(col_name)) %>% 
    rename(StnID = col_name) %>% 
    # select only the distinct StnIDs
    distinct() %>% 
    left_join(cus_wells) %>%
    select(!name) %>% 
    st_as_sf()
    
  plot <- ggplot() +
    # plot the outline of texas
    geom_sf(data = continental_us, fill = "white") +
    # plot all the wells in texas
    geom_sf(data = cus_wells, size = 0.05, color = "grey", alpha = 0.5) + 
    # plot the wells with breakpoints for that range
    geom_sf(data = plot_data, size = 0.05, color = "blue", alpha = 0.5) + 
    theme_classic()
  
  path = '/Users/lunacatalan/Documents/dev/Jasechko/groundwater-trends/plots'
  file_name = paste("plot_", col_name, ".png", sep="")
  ggsave(path = path, filename = file_name, width = 10, height = 5, dpi = 150, units = "in")
  
}

```

The next step:
1. identify locations (eg. regions or cities) that have blue clusters and record them so that they can be investigated further for groundwater level rise. 
2. The goal is to add more rows to the locations where groundwater levels may have increased. 
3. Doing it on a leaflet map could allow you to pan to specific locations and see if there are regional patterns - it could also add a level of reactivity

```{r, message=FALSE, warning = FALSE}

for (range in seq(cus_year_bins)) {
  
  range_df <- cus_year_bins[range]
  col_name <- colnames(range_df)
  
  plot_data <- range_df %>% 
    # unnest the list of station IDs
    unnest(cols = c(col_name)) %>% 
    rename(StnID = col_name) %>% 
    # select only the distinct StnIDs
    distinct() %>% 
    left_join(cus_wells) %>%
    select(!name) %>% 
    st_as_sf()
  
  # create leaflet map using the map data
  m <- leaflet() %>% 
    setView(-96, 37.8, 4) %>% 
    addTiles() %>% 
    addCircleMarkers(data = cus_wells, radius = 0.01, color = "grey") %>% 
    addCircleMarkers(data = plot_data,
                     radius = 0.01)
  
  # save the file using the date range naming conventions
  file_name = paste("plot_", col_name, ".html", sep="")
  saveWidget(m, file = file_name)
  
}

```


# Cities 

## 1953-1963
- Ciudad Juarez
- Artesia / Roswell 
- Carlsbad
- San Antonio
- New Braunfels
- Fort Collins
- Sedgwick/Burrton, Kansas

## 1964-1974
- San Antonio, Texas
- Fort Stockton, Texas
- Carlsbad, New Mexico
- Fillmore, Utah
- Santa Maria, California
- Hollister (Monterey), California 
- Hempstead, NY 
- Parsippany Troy Hilla, NJ
- Binghamton, NY

## 1975-1985
- Phoenix, Arizona
- Palmdate, CA
- Hollister (Monterey), California 
- Winnemucca
- Battle Mountain, NV
- Arco, ID
- Lehi, Utah
- Roswell, NM
- Houston, TX
- Elmer, Louisiana
- Baton Rouge, Louisiana
- Trinity, Florida 
- Aurora, Nebraska

## 1986-1996
- Phoenix, Arizona
- Hermiston, Oregon
- Howe, Idaho
- Lehi, Utah
- Ciudad Juarez
- Austin, TX
- Houston, TX
- Baton Rouge, Louisiana
- Albany, Georgia
- Valdosta, Georgia
- Trinity, Florida 
- Philadelphia, Pennsylvania 
- Hempstead, NY
- Grand Island, Nebraska
- Saint Paul, Minnesota 
- Monteview, Idaho




